Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-12 paper/main.tex
better spacing
compress section 7

\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}

@@

The rules are syntax directed and applied inductively.  For example,
the outermost expression is a loop, so we use the rule \text{Q:Loop}
at the root of the derivation tree.  At this point, we don't know yet
whether a loop invariant exists.  But we produce the constraints
$Q^\text{lo} = P^\text{lo}$\iffull{ which is short for the following
constraint set.
$$
\begin{array}{lll}
q^\text{lo}_0 = p^\text{lo}_0 &
q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
\end{array}
$$}{.}
%
This constraint expresses the fact that the potential before and
after the loop body is equal and thus constitutes an invariant.

In general, the weakening rule can be applied after every syntax directed
rule.  However, it can be left out in practice at some places to
increase the efficiency of the tool.
The weakening operation $\succeq_{\Gamma}$ is defined by the rule
\textsc{Relax}, it is parameterized by a logical context that is
used to gather information on interval sizes.  For example, we have
$$
\begin{array}{lll}
P^\text{de} \succeq_{(\cdot)} P^\text{we}
& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
\end{array}
$$
%
After the constraint generation, the LP solver is provided with an
objective function to minimize.  We wish to
minimize the initial potential, which constitutes the resource bound on the
whole program.  Here it is given by $Q$.  Moreover, we would like to
express that minimization of linear potential such as
$q_{10,x} |[10,x]|$
takes priority over minimization of constant potential such as
$q_{0,10} |[0,10]|$.

@@

\subsection{Cost Semantics for Clight}

To base the soundness proof on a formal ground, we start by defining a
new cost-aware operational semantics for Clight.  CompCert Clight's
operational semantics is based on small-step transitions and
continuations.  Expressions---which do not have side effects---are
evaluated in a big-step fashion.

@@

The small-step semantics we define is standard, except that it tracks
the resource consumption of a program.  The semantics is parametric in
the resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource \emph{metric}
$M$; a tuple of nine rational numbers and two maps.  Each of these
parameters indicates the amount of resource consumed by a corresponding
step in the semantics.  Resources can be released by assigning a negative
cost. %  The metric used in the implementation can also
% depend on other information that is statically available such as the
% name of the called function or the number of arguments.
Two rules for update and tick follow.
\begin{mathpar}
\Rule[rightskip=0.3cm]{U}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\rightarrow\\\\
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
\and
\Rule[leftskip=0.3cm,rightskip=0.2cm]{T}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\rightarrow\\\\
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}
\end{mathpar}
The rules keep track of the resource cost $\cost$ at all times, and
all rules have as implicit side condition that $\cost$ in the start
configuration is non-negative.  It makes it possible
to detect a resource crash as a stuck configuration where $\cost<0$.

@@

\subsection{Quantitative Hoare Logic}

\begin{figure*}
\small
\begin{mathpar}
%
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q + M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
\and \Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
\!\!  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  \!\! B' \models B \!\! \\
  \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
  \!\! P \models P' \!\! \\
  \!\! \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \!\! \\
  \!\! Q' \models Q \!\!
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Qplusz
}
{ \!\! \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x } \!\!
}
%
\end{mathpar}
\caption{Selected rules of the Quantitative Hoare Logic}
\label{fig:logic}
\end{figure*}

To prove the soundness of \toolname we found it useful to go through
an intermediate step using a Quantitative Hoare Logic.  This logic
is at the same time a convenient semantic tool and a clean way to
interface manual proofs with our automation, as explained in XXX.
The logic is an extension of the one presented in \cite{veristack14},
it adds support for arbitrary resources and simplifies the
handling of auxiliary state.

Following the previous work, we define quantitative Hoare triples
as $\Delta;B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
maps from program states to $\Qplusz\cup\{\infty\}$ in order to represent
an amount of resources available.  The assertions $B$ and $R$ are
postconditions for the case in which the block $S$ exits by a \code{break}
or \code{return} statement.  Additionally, $R$ depends on an extra value
that is returned by the function.  The meaning of the triple $\htriple Q S
{Q'}$ is as follows: If $S$ is executed with starting state $\state$,
the empty continuation \Kstop{} and at least $Q(\state)$ resource units
available then the evaluation does not run out of resources and there
are at least $Q(\state')$ resources left if the evaluation terminates
in state $\state'$.  The logic rules are similar to the ones in previous
work and are simply extended to account for all the costs found in
our cost-aware semantics.  Some rules are presented in \pref{fig:logic}.

Finally, we define a continuation based
soundness of triples and prove the soundness of all rules in Coq.  The
full version of this paper \cite{anon_extended}, gives more explanations
on the rules and provides a thorough overview of our soundness proof.

\subsection{The Soundness Theorem}

We use the
logic from previous paragraph as the target of a translation
function for the automatic derivation system.  This reveals two orthogonal
aspects of the proof: on one side, it relies on amortized reasoning
(the quantitative logic rules), and on the other side, it
uses combinatorial properties of our linear potential functions
(the automatic analysis rules).

>>>>

\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}
\jan{Here we have to say something about function specifications and
  function calls.}

@@

In general, the weakening rule can be applied after every syntax directed
rule.  However, it can be left out in practice at some places to
increase the efficiency of the tool.
The weakening operation $\succeq_{\Gamma}$ is defined by the rule
\textsc{Relax}, it is parameterized by a logical context that is
used to gather information on interval sizes.  For example, we have
$$
\begin{array}{lll}
P^\text{de} \succeq_{(\cdot)} P^\text{we}
& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
\end{array}
$$
%
The other rules are syntax directed and applied inductively.  For example,
the outermost expression is a loop, so we use the rule \text{Q:Loop}
at the root of the derivation tree.  At this point, we don't know yet
whether a loop invariant exists.  But we produce the constraints
$Q^\text{lo} = P^\text{lo}$\iffull{ which is short for the following
constraint set.
$$
\begin{array}{lll}
q^\text{lo}_0 = p^\text{lo}_0 &
q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
\end{array}
$$}{.}
%
This constraint expresses the fact that the potential before and
after the loop body is equal and thus constitutes an invariant.

After the constraint generation, the LP solver is provided with an
objective function to minimize.  We wish to
minimize the initial potential, which constitutes the resource bound on the
whole program.  Here it is given by $Q$.  Moreover, we would like to
express that minimization of linear potential such as
$q_{10,x} |[10,x]|$
takes priority over minimization of constant potential such as
$q_{0,10} |[0,10]|$.

@@

\paragraph{Cost Semantics for Clight.}
To base the soundness proof on a formal ground, we start by defining a
new cost-aware operational semantics for Clight.  CompCert Clight's
operational semantics is based on small-step transitions and
continuations.  Expressions---which do not have side effects---are
evaluated in a big-step fashion.

@@

The small-step semantics we define is standard, except that it tracks
the resource consumption of a program.  The semantics is parametric in
the resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource \emph{metric}
$M$; a tuple of nine rational numbers and two maps.  Each of these
parameters indicates the amount of resource consumed by a corresponding
step in the semantics.  Resources can be released by using a negative
cost. %  The metric used in the implementation can also
% depend on other information that is statically available such as the
% name of the called function or the number of arguments.
Two sample rules for update and tick follow.
\begin{mathpar}
\Rule[rightskip=0.3cm]{U}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\rightarrow\\\\
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
\and
\Rule[leftskip=0.3cm,rightskip=0.2cm]{T}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\rightarrow\\\\
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}
\end{mathpar}
The rules keep track of the resource cost $\cost$ at all times, and
all rules have as implicit side condition that $\cost$ in the start
configuration is non-negative.  It makes it possible
to detect a resource crash as a stuck configuration where $\cost<0$.

@@

% \begin{figure*}
% \small
% \begin{mathpar}
% %
% \Rule{L:Update}
% { }
% { \Delta; B; R \vdash
% \htriple
%   { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
% }
% %
% \and
% \Rule{L:Assert}
% { }
% { \Delta; B; R \vdash
% \htriple
%   { \code{istrue}~\sem{e}_\state \implies Q + M_a }
%   { \code{assert}~e }
%   { Q }
% }
% %
% \and
% \Rule{L:Tick}
% { }
% { \Delta; B; R \vdash
% \htriple
%   { Q + M_t(n) }
%   { \code{tick}(n) }
%   { Q }
% }
% %
% \and
% \Rule{L:Loop}
% { \Delta; Q; R \vdash
% \htriple
%   { I }{ S }{ I + M_l }
% }
% { \Delta; B; R \vdash
% \htriple
%   { I }{ \code{loop}~S }{ Q }
% }
% \and \Rule{L:Seq}
% { \Delta; B; R \vdash
% \htriple{ P }{ S_1 }{ Q' + M_s }
% \\
% \!\!  \Delta; B; R \vdash
% \htriple{ Q' }{ S_2 }{ Q }
% }
% { \Delta; B; R \vdash
% \htriple{ P }{ S_1;S_2 }{ Q }
% }
% %
% \and
% \Rule{L:Weaken}
% {
%   \!\! B' \models B \!\! \\
%   \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
%   \!\! P \models P' \!\! \\
%   \!\! \Delta; B'; R' \vdash
%   \htriple{ P' }{ S }{ Q' } \!\! \\
%   \!\! Q' \models Q \!\!
% }
% { \Delta; B; R \vdash
% \htriple{ P }{ S }{ Q }
% }
% %
% \and
% \Rule{L:Frame}
% { \Delta; B; R \vdash
% \htriple{ P }{ S }{ Q } \\
%   x \in \Qplusz
% }
% { \!\! \Delta; B + x; R + x \vdash
% \htriple{ P + x }{ S }{ Q + x } \!\!
% }
% %
% \end{mathpar}
% \caption{Selected rules of the Quantitative Hoare Logic}
% \label{fig:logic}
% \end{figure*}

\paragraph{Quantitative Hoare Logic.}
To prove the soundness of \toolname we found it useful to go through an
intermediate step using a Quantitative Hoare Logic.  This logic is at
the same time a convenient semantic tool and a clean way to interface
manual proofs with our automation.  We base it on the one presented in
\cite{veristack14}, we add support for arbitrary resources and simplify
the handling of auxiliary state.

Following this previous work, we define quantitative Hoare triples
as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
an amount of resources available.  The assertions $B$ and $R$ are
postconditions for the case in which the block $S$ exits by a \code{break}
or \code{return} statement.  Additionally, $R$ depends on the return
value of the current function.  The meaning of the triple $\htriple Q
S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
the empty continuation, and at least $Q(\state)$ resource units available
then the evaluation does not run out of resources and there are at
least $Q(\state')$ resources left if the evaluation terminates in
state $\state'$.  The logic rules are similar to the ones in previous
work and are simply extended to account for all the costs found in our
cost-aware semantics.

Finally, we define a strong compositional continuation based soundness for
triples and prove the validity of all the rules in Coq.  \ifshort{The full
version of this paper \cite{anon_extended}, provides explanations for
many rules and a thorough overview of our soundness proof.}{Details about
the logic and its soundness proof are provided in the \pref{app:logic}.}

\paragraph{The Soundness Theorem.}
We use the logic from previous paragraph as the target of a translation
function for the automatic derivation system.  This reveals two orthogonal
aspects of the proof: on one side, it relies on amortized reasoning (the
quantitative logic rules), and on the other side, it uses combinatorial
properties of our linear potential functions (the automatic analysis
rules).
